---
title: "CalCOFI Station Portal ↔ Integrated Database — Crosswalk Report"
subtitle: "Which portal datasets & variables are matching, partial, missing, or new"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    theme: cosmo
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
librarian::shelf(readr, dplyr, tidyr, DT, glue, quiet = TRUE)

xd <- read_csv("crosswalk_datasets.csv",  show_col_types = FALSE)
xv <- read_csv("crosswalk_variables.csv", show_col_types = FALSE)

# consistent class ordering + colors
class_lvls <- c("matching", "partial", "missing", "new")
class_pal  <- c(matching = "#2f9e44", partial = "#f08c00",
                missing = "#e03131", new = "#1971c2")

dt <- function(d, ...) datatable(
  d, rownames = FALSE, filter = "top", extensions = "Buttons",
  options = list(pageLength = 15, scrollX = TRUE, dom = "Bfrtip",
                 buttons = c("copy", "csv")), ...)
```

This report reconciles the **existing station portal** (its `data_sources.csv`
datasets and harvested `variables.json`) against the **CalCOFI integrated
database** (`provider_dataset` + `measurement_type`/taxa). Names differ between
the two, so each entry is classed:

- **matching** — portal entry has a direct integrated-DB counterpart
- **partial** — overlapping but a different provider slice / grain
- **missing** — portal has it, the DB does *not* (→ ingest backlog)
- **new** — the DB has it, the portal does *not* yet surface it

Source tables: `crosswalk_datasets.csv`, `crosswalk_variables.csv` (regenerate
the latter with `duckdb -c ".read scripts/build_crosswalk.sql"`).

# Datasets

```{r ds-topline}
ds_by_class <- xd |>
  mutate(match_class = factor(match_class, class_lvls)) |>
  count(match_class, .drop = FALSE) |>
  mutate(pct = round(100 * n / sum(n)))
```

**`r nrow(xd)` dataset rows** — `r sum(xd$match_class=='matching')` matching ·
`r sum(xd$match_class=='partial')` partial ·
`r sum(xd$match_class=='missing')` missing ·
`r sum(xd$match_class=='new')` new.

```{r ds-by-class}
ds_by_class |>
  transmute(`match class` = match_class, count = n, `%` = pct) |>
  dt(caption = "Dataset rows by match class")
```

```{r ds-by-realm}
xd |> count(realm, match_class) |>
  pivot_wider(names_from = match_class, values_from = n, values_fill = 0) |>
  dt(caption = "Dataset match class × realm")
```

## Full dataset crosswalk

```{r ds-full}
xd |>
  select(portal_dataset_id, portal_name, realm, db_provider_dataset, match_class, notes) |>
  dt(caption = "Every portal dataset mapped to the integrated DB") |>
  formatStyle("match_class", target = "cell",
    backgroundColor = styleEqual(names(class_pal), unname(class_pal)),
    color = "white", fontWeight = "bold")
```

# Variables

```{r var-topline}
n_env  <- sum(grepl("measurement_type", xv$var_note))
n_tax  <- sum(grepl("taxon", xv$var_note))
n_unm  <- sum(xv$var_note == "unmatched")
n_nod  <- sum(grepl("no DB dataset", xv$var_note))
```

**`r nrow(xv)` variable rows** — `r n_env` matched to a DB `measurement_type`,
`r n_tax` classed as taxa (resolved at build time by species name), `r n_unm`
unmatched (mostly identifier/QC columns), `r n_nod` with no DB dataset
(missing/omics).

```{r var-by-note}
xv |> count(var_note, sort = TRUE) |>
  dt(caption = "Variable rows by crosswalk outcome")
```

```{r var-by-dataset}
xv |>
  mutate(dataset_match_class = factor(dataset_match_class, class_lvls)) |>
  count(db_provider_dataset, dataset_match_class, .drop = FALSE) |>
  filter(!is.na(db_provider_dataset)) |>
  pivot_wider(names_from = dataset_match_class, values_from = n, values_fill = 0) |>
  dt(caption = "Variables per integrated-DB dataset, by the dataset's match class")
```

## Full variable crosswalk

```{r var-full}
xv |>
  select(portal_dataset_id, variable_name, display_name, units, entity_type,
         db_provider_dataset, realm, dataset_match_class, measurement_type_match, var_note) |>
  dt(caption = "Every portal variable mapped to the integrated DB")
```

# Insights & recommendations

```{r gaps}
missing_ds <- xd |> filter(match_class == "missing") |> pull(portal_name)
new_ds     <- xd |> filter(match_class == "new")     |> pull(db_provider_dataset)
partial_ds <- xd |> filter(match_class == "partial") |> pull(portal_name)
```

**Coverage is strong for the core program.** Every headline environmental and
biological program the portal advertises resolves to an integrated-DB dataset:
hydrographic bottle → `calcofi_bottle`, CTD → `calcofi_ctd-cast`, carbonate
chemistry → `calcofi_dic`, the six ERDDAP ichthyoplankton tables →
`swfsc_ichthyo`, CUFES → `swfsc_cufes`, zooplankton biovolume →
`pic_zooplankton`, euphausiids → `cce-lter_euphausiids`, ZooDB → `cce-lter_zoodb`,
phytoplankton → `calcofi_phytoplankton`.

**Ingest backlog (portal advertises, DB lacks — `missing`):**
`r glue::glue_collapse(missing_ds, sep = "; ")`. These are candidate future
ingests (underway meteorology, microbial/picoplankton, Dungeness crab
postlarvae, the Stanford historical zooplankton archive, mesopelagic fish, and
two eDNA/genomics ASV datasets — genomics is out of scope for the current
observation model).

**Surface in the portal (DB has, portal lacks — `new`):**
`r glue::glue_collapse(new_ds, sep = "; ")`. `calcofi_phyllosoma` and
`cce-lter_zooscan` are fully ingested but not yet advertised — add them to
`data_sources.csv` / the map legend.

**Partial mappings to review (`partial`):**
`r glue::glue_collapse(partial_ds, sep = "; ")` all fold into
`calcofi_bird_mammal_census`, but come from different provider slices (Farallon
Institute seabirds; CCE-LTER underway mammal sightings) than the CalCOFI census
— confirm these are the same underlying observations before merging.

**Data-quality fixes surfaced by the crosswalk:**

- `calcofi_phytoplankton` was present in `measurement_type.csv` and had a full
  ingest, but was **missing from `metadata/dataset.csv`** — now added.
- **Euphausiids lack per-species resolution** in the integrated DB (only total
  `euphausiid_abundance`); the portal's ~39 harvested krill species are not
  backed by DB data. Re-ingest with the species dimension to close this.
- The **taxon → harvested-extras** join is currently name-based and low-yield;
  a future pass should match on WoRMS `aphia_id` (the DB taxa carry it) rather
  than scientific-name strings.

**Bottom line:** the portal can be driven entirely by the integrated DB today
for the core datasets, with a short, explicit backlog (above) rather than silent
gaps.
```
